---
title: "Health care fraud detection"
author: "Spandana Makeneni"
date: "10/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading libraries and input files 
```{r }
library(tidyverse)
library(ggthemr)
library(ggpubr)
library(naniar)
library(ggridges)
library(moments)
library(caret)
library(rcompanion)
library(gbm)
library(xgboost)
library(Metrics)
ggthemr(palette = "solarized")

train_beneficiary <- read.csv("/Users/spandana/Desktop/Data_science_projects/Healthcare_fraud_detection/Train_Beneficiarydata.csv")
train_inpatient <- read.csv("/Users/spandana/Desktop/Data_science_projects/Healthcare_fraud_detection/Train_Inpatientdata.csv")
train_outpatient <- read.csv("/Users/spandana/Desktop/Data_science_projects/Healthcare_fraud_detection/Train_Outpatientdata.csv")
train_labels <- read.csv("/Users/spandana/Desktop/Data_science_projects/Healthcare_fraud_detection/Train.csv")
States <- read.csv("/Users/spandana/Desktop/Data_science_projects/Healthcare_fraud_detection/States.csv",header=FALSE)
colnames(States)<-c("State","State_name")
```

#Data sets

**Inpatient Dataset** - This data provides insights about the claims filed for those patients who are admitted in the hospitals. It also provides additional details like their admission and discharge dates and admit d diagnosis code.

**Outpatient Dataset** - This data provides details about the claims filed for those patients who visit hospitals and not admitted in it.

**Beneficiary Dataset** - This data contains beneficiary KYC details like health conditions,region they belong to etc.

**Lets take a look at the dimensions and datatypes of each dataset**

```{r}
print("Column names and data types for Inpatient Dataset")
str(train_inpatient)
print("Column names and data types for Outpatient Dataset")
str(train_outpatient)
print("Column names and data types for Beneficiary Dataset")
str(train_beneficiary)
```

Insights:
1) Dates were read as factors. Lets change that.
2) There are some missing values. Lets take a look at those.

```{r formattingdates}
date_colnames <- c("ClaimStartDt","ClaimEndDt","AdmissionDt","DischargeDt","DOB","DOD")
train_inpatient[,date_colnames[1:4]] <- lapply(train_inpatient[,date_colnames[1:4]],as.Date)
train_outpatient[,date_colnames[1:2]] <- lapply(train_outpatient[,date_colnames[1:2]],as.Date)
train_beneficiary[,date_colnames[5:6]] <- lapply(train_beneficiary[,date_colnames[5:6]],as.Date)
```

#Data preprocessing 
#Missing values
```{r checkingmissingvalues}

#Function to plot missing values 
plot_missingvalues <- function(df,title_name){
  df_missing <- miss_var_summary(df)
  p <- ggplot(df_missing,aes(x=reorder(variable,-pct_miss),y=pct_miss,label=round(pct_miss,1)))+geom_segment(aes(x=reorder(variable,-pct_miss),xend=reorder(variable,-pct_miss),y=0,yend=pct_miss))+geom_point(size=4,alpha=0.6)+geom_text(nudge_x =0.2,nudge_y=3.5)+labs(y="% missing",x="",title=title_name)+coord_flip()
  print(p)
}

plot_missingvalues(train_inpatient,"Missing values in Inpatient data")
plot_missingvalues(train_outpatient,"Missing values in Outpatient data")
plot_missingvalues(train_beneficiary,"Missing values in Beneficiary data")
```
#Data cleaning 

Step1:   
Dropping the following columns because 100% of the data is missing:  
1) ClmProcedure code 4,5,6 from the inpatient data 
2) ClmProcedureCode 1,2,3,4,5,6 from the outpaitent data -  After researching a little I found that frequently these columns exist in the claims data but when they are not the basis of payment they are empty. 
3) DOD - Date of Death  

```{r dropping columns}
train_inpatient <- train_inpatient %>% select(-ClmProcedureCode_4,-ClmProcedureCode_5,-ClmProcedureCode_6)
train_outpatient <- train_outpatient %>%  select(-starts_with("ClmProcedure"))
train_beneficiary <- train_beneficiary %>% select(-DOD)
```

Ok, lets deal with the remaining missing values in each data set  

Step 2:  
Inpatient and Outpatient data sets- Missing values in Physician columns, Diagnosis codes, and procedure codes.

For Physician columns - Replace NAs with "None"
For Diagnosis codes  and procedure codes - Replace NAs with 0 
```{r replotting missing values after deleting some columns}
replace_NAs_None <- function(df){
df$AttendingPhysician <- factor(df$AttendingPhysician,levels=c(levels(df$AttendingPhysician),"None"))
df$OperatingPhysician <- factor(df$OperatingPhysician,levels=c(levels(df$OperatingPhysician),"None"))
df$OtherPhysician <- factor(df$OtherPhysician,levels=c(levels(df$df$OtherPhysician),"None"))
df$AttendingPhysician[is.na(df$AttendingPhysician)] <- "None"
df$OperatingPhysician[is.na(df$OperatingPhysician)] <- "None"
df$OtherPhysician[is.na(df$OtherPhysician)] <- "None"
return(df)
}

train_inpatient <- replace_NAs_None(train_inpatient)
train_inpatient <- train_inpatient %>% mutate_if(is.factor,fct_explicit_na,na_level="0")
train_inpatient[is.na(train_inpatient)] <-0 
train_outpatient <- replace_NAs_None(train_outpatient)
train_outpatient <- train_outpatient %>% mutate_if(is.factor,fct_explicit_na,na_level="0")
train_outpatient[is.na(train_outpatient)] <- 0
```

Alright,there are no missing values.

Step 3:
Chronic conditions are listed as 1=Yes, 2=No
We will convert 2 to 0 just for ease of computation 
Add a new column age will be more useful than date of birth
Adding a new column of state names 

```{r converting 2s to 0s}
train_beneficiary[,10:20][train_beneficiary[,10:20]==2] <- 0
train_beneficiary <- train_beneficiary %>% mutate(Age=round(as.numeric(as.Date("2009-12-31")-DOB)/365))
train_beneficiary <- left_join(train_beneficiary,States)
```

Step 4:
Add a column for total duration in hospital =  DischargeDt-AdmissionDt
```{r adding duration}
train_inpatient <- train_inpatient %>% mutate(Totalstay_days=round(as.numeric(DischargeDt-AdmissionDt))+1)
train_inpatient <- train_inpatient %>% mutate(Claimlength=round(as.numeric(ClaimEndDt-ClaimStartDt)))
train_inpatient <- train_inpatient %>% mutate(Type="Inpatient")

train_outpatient <- train_outpatient %>% mutate(Claimlength=round(as.numeric(ClaimEndDt-ClaimStartDt)))
train_outpatient <- train_outpatient %>% mutate(Type="Outpatient")
```

#Joining the beneficiary and Inpatient, Outpatient datsets for analysis 

```{r joining datasets}
Merged_in_ben_fruad<- inner_join(train_inpatient,train_beneficiary)
Merged_in_ben_fruad<- inner_join(Merged_in_ben_fruad,train_labels)

Merged_out_ben_fraud <- inner_join(train_outpatient,train_beneficiary)
Merged_out_ben_fraud <- inner_join(Merged_out_ben_fraud,train_labels)

all_data<-bind_rows(Merged_in_ben_fruad,Merged_out_ben_fraud)
all_data<-all_data %>% mutate_if(is.factor,fct_explicit_na,na_level="0")
all_data$AdmissionDt[is.na(all_data$AdmissionDt)] <- as.Date("1970-01-01")
all_data$DischargeDt[is.na(all_data$DischargeDt)] <- as.Date("1970-01-01")
all_data[is.na(all_data)] <-0
all_data$Gender[all_data$Gender==2]<-0
all_data <- all_data %>% select(-DOB,-State)
```

Ok, lets explore the data now.

#Exploratory data analysis

First, lets take a look at the target variable 

```{r plotting the target variable}
ggplot(all_data,aes(x=factor(PotentialFraud)))+geom_bar()+geom_text(stat="count",aes(label=paste(round(stat(prop)*100,0),"%"),group=1,vjust=2.0))+facet_wrap(~Type,scales="free_y")+ labs(y="total",x="",title="Fraud claims in Inpatient and Outpatient data")

ggplot(all_data,aes(x=factor(PotentialFraud)))+geom_bar()+geom_text(stat="count",aes(label=paste(round(stat(prop)*100,0),"%"),group=1,vjust=2.0))


```
Insights: 
Data sets are slightly imbalanced - What are the ways to deal with this?
Inpatient has more fraud than outpatient

#Age and Fraud
```{r}
#Age distribution
#Adding Age bin
all_data <- all_data %>% mutate(Agebin=ifelse(Age>=20 & Age<=40,"20-40",
                            ifelse(Age>40 & Age<=60,"40-60",
                            ifelse(Age>60 & Age<=80,"60-80",
                            ifelse(Age>80 & Age<=100,"80-100","None")))))

ggplot(all_data,aes(x=Age))+geom_histogram(aes(y=..density..),colour="black",fill="white")+geom_density(alpha=0.2)
ggplot (all_data,aes(x=Age,color=PotentialFraud,fill=PotentialFraud))+geom_histogram(aes(y=..density..),alpha=0.5,position="identity")+geom_density(alpha=0.2)

#Age vs Fraud
ggplot(all_data,aes(x=PotentialFraud,y=Age,fill=PotentialFraud))+geom_boxplot()
ggplot(all_data,aes(x=Age,y=PotentialFraud,fill=PotentialFraud))+geom_density_ridges(alpha=0.6,stat="binline",bins=20)

```
Insight:
Age seems to have no impact on Fraud

#Gender and Fraud
```{r}
ggplot(all_data,aes(x=factor(Gender),fill=PotentialFraud))+geom_bar(position=position_dodge())
#ggplot(all_data,aes(x=factor(PotentialFraud),fill=factor(Gender)))+geom_bar(position = position_dodge())+
 # geom_text(stat="count",aes(label=scales::percent(..prop..),group=1))
```
Insights:
There are more females in the dataset than males
No influence on fraud

#Race and Fraud
```{r }
ggplot(all_data,aes(x=Race,fill=PotentialFraud))+geom_bar(position = position_dodge())
```

Insights:
Very imbalanced - maybe delete it 

#State and Fraud 
```{r }
state_data <- all_data %>% group_by(State_name,PotentialFraud) %>% summarize(total=n())
state_data <- spread(state_data,PotentialFraud,total)
state_data <- state_data[-c(1),]
colnames(state_data) <- c("region","No","Yes")
state_data$region <- tolower(state_data$region)
states_map <- map_data("state")
map_state_date <- left_join(states_map,state_data,by="region")
state_abbs <- as.data.frame(cbind(state.abb,state.name))
colnames(state_abbs) <- c("state_abb","region")
state_abbs$region <- tolower(state_abbs$region)
map_state_date <- left_join(map_state_date,state_abbs,by="region")
state_names <- map_state_date %>% group_by(state_abb) %>% summarize(lat=0.5*(max(lat)+min(lat)),long=0.5*(max(long)+min(long)))
map_state_date <- map_state_date %>% mutate(total=No+Yes)

ggplot(map_state_date,aes(x=long,y=lat))+geom_polygon(aes(group=group,fill=Yes),color="black")+scale_fill_continuous(low="lightblue",high="darkblue")+geom_text(data=state_names,mapping=aes(x=long,y=lat,label=state_abb),color="white",size=3)

ggplot(map_state_date,aes(x=long,y=lat))+geom_polygon(aes(group=group,fill=total),color="black")+scale_fill_continuous(low="lightblue",high="darkblue")+geom_text(data=state_names,mapping=aes(x=long,y=lat,label=state_abb),color="white",size=3)

state_test <- all_data %>% select(Provider,State_name,PotentialFraud) %>% count(Provider,State_name,PotentialFraud) %>% spread(State_name,n) 
state_test[is.na(state_test)] <-0
state_test$total <- apply(state_test[,4:54],1,function(x)sum(x>0))
ggplot(state_test,aes(total))+geom_histogram()+facet_wrap(~PotentialFraud,scales="free_y")

#state_data %>% select(region,Yes) %>% arrange(desc(Yes))

```

#Looks like California, Florida, Newyork, PA, Texas have higher fraud claims than the rest of country.

#Chronic conditions and Fraud 
```{r }
all_data$Total_chronicconds <- as.numeric(apply(all_data[,37:47],1,sum))
all_data <- all_data %>% mutate(chroniccond=ifelse(Total_chronicconds>0,"yes","no"))
ggplot(all_data,aes(x=Total_chronicconds))+geom_histogram()+facet_wrap(~PotentialFraud,scales="free_y")
ggplot(all_data,aes(Age,Total_chronicconds))+geom_bar(stat="identity")+facet_wrap(~PotentialFraud,scales="free_y")
ggplot(all_data,aes(x=as.factor(PotentialFraud),y=Total_chronicconds))+geom_bar(stat="identity",position=position_dodge())
cc <- all_data %>% select(starts_with("Chronic"),PotentialFraud) %>% gather(chroniccond,value,starts_with("Chronic"))
cc %>% group_by(chroniccond,value,PotentialFraud) %>% summarize(total=n()) %>% ggplot(aes(x=reorder(chroniccond,total),y=total,fill=PotentialFraud))+geom_bar(stat="identity",position=position_dodge())+theme(axis.text.x = element_text(angle=90))
cc

```

#Diagnosis codes and Fraud 
```{r topdiagnosiscodesassociatedwithfraud}
Top_diagnosiscodes <- all_data %>% select(starts_with("ClmDiagnosisCode"),PotentialFraud,Type) %>% gather(Code,Value,starts_with("ClmDiagnosisCode")) %>% filter(PotentialFraud=="Yes")

Top_diagnosiscodes_nf <- all_data %>% select(starts_with("ClmDiagnosisCode"),PotentialFraud,Type) %>% gather(Code,Value,starts_with("ClmDiagnosisCode")) %>% filter(PotentialFraud=="No")

Overall_top_diagnosiscodes <- Top_diagnosiscodes %>% filter(Value!=0) %>% group_by(Value) %>% summarize(total=n()) %>% arrange(desc(total)) 

Overall_top_diagnosiscodes_nf <- Top_diagnosiscodes_nf %>% filter(Value!=0) %>% group_by(Value) %>% summarize(total=n()) %>% arrange(desc(total)) 

#plotting the top ten codes
Overall_top_diagnosiscodes[1:10,] %>% ggplot(aes(x=reorder(Value,total),y=total))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=80))

#Adding percent column
#Overall_top_diagnosiscodes <- Overall_top_diagnosiscodes %>% mutate(percent_fraud=total/212796)
#Overall_top_diagnosiscodes_nf<- Overall_top_diagnosiscodes_nf %>% mutate(percent_notfraud=total/345415)
all_diagnosis_codes <- inner_join(Overall_top_diagnosiscodes,Overall_top_diagnosiscodes_nf,by="Value")
all_diagnosis_codes <- all_diagnosis_codes %>% mutate(percent_fraud=total.x/212796)
all_diagnosis_codes <- all_diagnosis_codes %>% mutate(percent_notfraud=total.y/345415)

ggplot(all_diagnosis_codes,aes(percent_fraud,percent_notfraud))+geom_point()+scale_x_log10()+scale_y_log10()+geom_abline(color="black")

```
Looks like some codes are strongly associated with fraudulent claims 

```{r top diangnosis codes not associated with fraud}
#Top_diagnosiscodes_nf <- all_data %>% select(starts_with("ClmDiagnosisCode"),PotentialFraud,Type) %>% gather(Code,Value,starts_with("ClmDiagnosisCode")) %>% filter(PotentialFraud=="No")
#Top_diagnosiscodes_group_nf <- Top_diagnosiscodes_nf %>% group_by(Code,Value) %>% summarize(total=n()) %>% arrange(desc(total)) %>% slice(1:10)
#Overall_top_diagnosiscodes_nf <- Top_diagnosiscodes_nf %>% filter(Value!=0) %>% group_by(Value) %>% summarize(total=n()) %>% arrange(desc(total)) %>% slice(1:10)
#Top_diagnosiscodes_group_nf %>% filter(Value!="0") %>% ggplot(aes(reorder(Value,total),total))+geom_bar(stat="identity")+facet_wrap(~Code,scales="free_x")+theme(axis.text.x = element_text(angle=80))
#Overall_top_diagnosiscodes_nf %>% ggplot(aes(x=reorder(Value,total),y=total))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=80))
```

#Procedure codes vs Fraud 
```{r }
Top_procedurecodes <- all_data %>% select(starts_with("ClmProcedureCode"),PotentialFraud,Type) %>% gather(Code,Value,starts_with("ClmProcedureCode")) %>% filter(PotentialFraud=="Yes")

Top_procedurecodes_nf <- all_data %>% select(starts_with("ClmProcedureCode"),PotentialFraud,Type) %>% gather(Code,Value,starts_with("ClmProcedureCode")) %>% filter(PotentialFraud=="No")

#Top_procedurecodes_group <- Top_procedurecodes %>% group_by(Code,Value) %>% summarize(total=n()) %>% arrange(desc(total)) %>% slice(1:10)

Overall_top_procedurecodes <- Top_procedurecodes %>% filter(Value!=0) %>% group_by(Value) %>% summarize(total=n()) %>% arrange(desc(total)) 
Overall_top_procedurecodes_nf <- Top_procedurecodes_nf %>% filter(Value!=0) %>% group_by(Value) %>% summarize(total=n()) %>% arrange(desc(total))
all_procedure_codes <- inner_join(Overall_top_procedurecodes,Overall_top_procedurecodes_nf,by="Value")
all_procedure_codes <- all_procedure_codes %>% mutate(percent_fraud = total.x/212796)
all_procedure_codes <- all_procedure_codes %>% mutate(percent_notfraud = total.y/345415)

ggplot(all_procedure_codes,aes(percent_fraud,percent_notfraud))+geom_point()+scale_x_log10()+scale_y_log10()+geom_abline(color="black")

Overall_top_procedurecodes[1:10,] %>% ggplot(aes(x=reorder(Value,total),y=total))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=80))
```
#Hospital stay and fraud 
```{r }
all_data %>% count(Totalstay_days,PotentialFraud) %>%  arrange(desc(n)) %>% filter(Totalstay_days!=0) %>% ggplot(aes(reorder(Totalstay_days,Totalstay_days),n))+geom_bar(stat="identity")+facet_wrap(~PotentialFraud)

```
#Claim length and fraud 
```{r }
all_data %>% count(Claimlength,PotentialFraud) %>% filter(Claimlength!=0) %>% arrange(desc(n)) %>%  ggplot(aes(Claimlength,n))+geom_bar(stat="identity")+facet_wrap(~PotentialFraud)

```
#Number of claims and Fraud 
```{r}
Totalclaims_byuser <- all_data %>% group_by(BeneID) %>% summarize(total=n())
Fraud_byuser <- all_data %>%count(BeneID,PotentialFraud) 
Fraud_byuser <- Fraud_byuser %>% spread(PotentialFraud,n)
Fraud_byuser[is.na(Fraud_byuser)]<-0
Totalclaims_byuser <- inner_join(Totalclaims_byuser,Fraud_byuser)
Totalclaims_byuser <- Totalclaims_byuser %>% mutate(Fraud = ifelse(Yes>0,"yes","no"))
ggplot(Totalclaims_byuser,aes(x=total,fill=Fraud))+geom_histogram( color="#e9ecef", alpha=0.6, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080"))+labs(fill="")+facet_wrap(~Fraud,scales="free_y")

```
#Total number of codes and Fraud 
```{r }
all_data$total_diagnosiscodes <- apply(all_data[,15:24],1,function(x)sum(x>0))
all_data$total_procedurecodes <- apply(all_data[,25:27],1,function(x)sum(x>0))
ggplot(all_data,aes(x=total_diagnosiscodes))+geom_histogram(color="#e9ecef",binwidth=1,alpha=0.6,position="identity")+scale_fill_manual(values=c("#69b3a2", "#404080"))+facet_wrap(~PotentialFraud,scales="free_y")
ggplot(all_data,aes(x=total_procedurecodes,fill=PotentialFraud))+geom_histogram(color="#e9ecef",binwidth=1,alpha=0.6,position="identity")+scale_fill_manual(values=c("#69b3a2", "#404080"))

```

#Getting the dataset ready 
```{r}
top_codes_list <- c("4280","2720","2449","V5861","4011","42731","V5869","2724","25000","4019")
codes_data <- all_data %>% select(starts_with("ClmDiagnosisCode"),Provider) %>% gather(Code,codevalue,starts_with("ClmDiagnosisCode")) %>% group_by(Provider,codevalue) %>% summarize(total=n()) %>% filter(codevalue %in% top_codes_list)
codes_data <- spread(codes_data,codevalue,total)
Age_data <- all_data %>% select(Agebin,Provider) %>% gather(Age_bin,bin_value,Agebin) %>% group_by(Provider,bin_value) %>% summarize(total=n()) %>% spread(bin_value,total)
Remaining_features <- all_data %>% group_by(Provider) %>% summarise(Totalstay_days=mean(Totalstay_days),total_diagnosiscodes=mean(total_diagnosiscodes),Total_chronicconds=mean(Total_chronicconds))


modeling_dataframe <- all_data %>% select(Provider,BeneID) %>% group_by(Provider) %>% mutate(Total_patients=n_distinct(BeneID))
modeling_dataframe <- modeling_dataframe %>% group_by(Provider,Total_patients) %>% summarize(Total_claims=n())
modeling_dataframe <- full_join(modeling_dataframe,codes_data)
modeling_dataframe <- full_join(modeling_dataframe,Age_data)
modeling_dataframe <- full_join(modeling_dataframe,state_test[,c(1,55)])
modeling_dataframe <- full_join(modeling_dataframe,Remaining_features)
colnames(modeling_dataframe)[colnames(modeling_dataframe)=="total"] <-"Total_states"
modeling_dataframe <- modeling_dataframe %>% select(-None)
modeling_dataframe[is.na(modeling_dataframe)]<-0
modeling_dataframe <- full_join(modeling_dataframe,train_labels)

```

#Scaling and Transforming
```{r }

#Function to plot skewness and skew values 
distbn_plot <- function(df,xname){
  skew<-skewness(df[,xname])
  #bw=(max(df[,xname])-min(df[,xname]))/(17)
  p <- ggplot(df,aes_string(xname))+
  geom_histogram( )+labs(title=paste("skew: ",round(skew,2)))
  print(p)
}  

for(i in colnames(modeling_dataframe[2:21])){
  print(i)
   distbn_plot(modeling_dataframe,i)
}

skewed <- sapply(modeling_dataframe[,2:21],function(x) {ifelse(abs(skewness(x)) > .8,"Yes","No")})
skewed_colnames <- names(skewed[skewed=="Yes"])
skewed_colnames

#tranforming data
#Normalizing skewed numerical columns
#for(col in skewed_colnames)
#{
#  print(col)
#  modeling_dataframe[,col] <- transformTukey(modeling_dataframe[,col])
#}

#Scaling variables
scaled_vals <- preProcess(modeling_dataframe[,2:21], method=c("center", "scale"))
modeling_dataframe[,2:21] <- predict(scaled_vals,modeling_dataframe[,2:21])

#Model
modeling_dataframe <- as.data.frame(modeling_dataframe)
set.seed(143)
train_partition <- createDataPartition(modeling_dataframe$PotentialFraud,p=0.7,list=F)
train_model <- modeling_dataframe[train_partition,]
validate_model <- modeling_dataframe[-train_partition,]
```


#Modeling
```{r}
logisticmodel <- glm(PotentialFraud ~ .,data=train_model,family="binomial"(link="logit"))
summary(logisticmodel)
```

1) Logistic Regression


Relationship between 
- Age and Fraud - Done 
- Gender and Fraud - Done
- Race and fraud - Done 
- Chronicconditions and Fraud -Done
- State and Fraud- Done 
- Diagnosis codes and Fraud - Done
- Procedure codes and Fraud - Done
- Hospital stay length and Fraud -Done 
- Claim length and Fraud -Done 
- Number of claims and Fraud - Have to figure this out 
- Inpatient vs Outpatient - Done 




